### 2021-06-17 submission/manuscript.tex
Additional edits

Diarrheal samples from patients taking laxatives are typically rejected
for \emph{Clostridiodes difficile} testing. However, there are
similarities between the bacterial communities from people with diarrhea
or \emph{C. difficile} infections (CDI) including lower diversity
compared to communities from healthy patients, which led us to
hypothesize that diarrhea may be an indicator of \emph{C. difficile}
risk. We explored how osmotic laxatives disrupt the microbiota's
colonization resistance to \emph{C. difficile} by administering a
laxative to mice either before or after \emph{C. difficile} challenge.
Our findings suggest that osmotic laxatives disrupt colonization
resistance to \emph{C. difficile}, and prevent clearance among mice
already colonized with \emph{C. difficile}. Considering that most
hospitals recommend not performing \emph{C. difficile} testing on
patients taking laxatives and laxatives are used when administering
fecal microbiota transplants via colonoscopy to patients with recurrent
CDIs, further studies are needed to evaluate if laxatives impact
microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative samples from patients are
separated into two groups based on stool consistency, there are similar
microbiota features between samples from CDI patients and \emph{C.
difficile} negative patients with diarrhea compared to samples that were
\emph{C. difficile} negative with non-diarrheal stool consistency (12).
The similar community features between CDI patients and patients with
diarrhea included low alpha diversity and only 6 bacterial taxa had
higher relative abundances in communities from CDI patients. These
results led to the hypothesis that bacterial communities from patients
experiencing diarrhea are susceptible to developing CDIs, regardless of
how they developed diarrhea.

depending on the dose administered, osmotic laxatives can lead to
diarrhea and temporarily disrupt the human intestinal microbiota (13).
The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
in Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} colonization
(14--17). The mucus barrier is thought to mediate protection from CDIs
by protecting intestinal epithelial cells from the toxins produced by
\emph{C. difficile} (18, 19). Whether laxative administration results in
more severe CDIs in mice and how long mice remain colonized with
\emph{C. difficile} after challenge is unclear.

@@

Because time and treatment group influenced most of the variation
between communities, we next profiled community diversity and
composition. We examined the alpha diversity dynamics by calculating the
communities' Shannon diversity. Although both clindamycin and PEG
treatments decreased diversity, the Shannon diversity was lower in the
groups of mice that received PEG treatment compared to those that
received clindamycin alone through thirty days post-challenge (Fig. 2C;
Data Set S1, sheet 2). We next identified the bacterial genera whose
relative abundances shifted after PEG treatment by comparing the
baseline samples of mice treated with only PEG to samples from the same
mice one day post-PEG-treatment. We found 18 genera whose relative
abundances were altered by PEG treatment (Data Set S1, sheet 3). The
majority of the bacterial relative abundances decreased after the PEG
treatment, but the relative abundance among members of the
\emph{Enterobacteriaceae} and \emph{Bacteroides} increased. The increase
in \emph{Bacteroides} relative abundance was unique to PEG treated mice,
as the \emph{Bacteroides} relative abundance actually decreased in
clindamycin treated mice (Fig. 2D). Finally, we identified the genera
whose relative abundance differed across treatment groups over multiple
time points. Of the 33 genera that were different between treatment
groups, 24 genera were different over multiple time points (Fig. 2E,
Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal
microbiota that was maintained over time and was distinct from
clindamycin treatment.

@@

One possible explanation for the prolonged \emph{C. difficile}
colonization in 5-day PEG treated mice, might be due to the bacteria's
persistence in the mucosal compartment. In fact, it has been
hypothesized that \emph{C. difficile} biofilms may serve as reservoirs
for recurrent infections (35) and \emph{C. difficile} biofilms in the
mucus layer were recently identified in patients as aggregates with
\emph{Fusobacterium nucleatum} (36). There was an interesting pattern of
increased \emph{Enterobacteriaceae}, \emph{Bacteroides}, and \emph{C.
difficile} in both the stool and mucosal communities of PEG-treated mice
suggesting a potential synergy. \emph{Bacteroides} has the potential to
degrade mucus and the osmotic laxative may have allowed
\emph{Bacteroides} to colonize the mucosal niche by degrading mucin
glycans with glycosyl hydrolases that are absent in \emph{C. difficile}
(37). \emph{Bacteroides} persistent in the mucosal tissue might also
have helped \emph{Enterobacteriaceae} to colonize the region, as a
synergy between mucus-degrading \emph{B. fragilis} and \emph{E. coli}
has previously been described (38). A separate study demonstrated
\emph{C. difficile} was present in the outer mucus layer and associated
with \emph{Enterobacteriaceae} and \emph{Bacteroidaceae} using
fluorescent in situ hybridization (FISH) staining (39). However,
protective roles for \emph{Bacteroides} have also been demonstrated. For
example, \emph{B. fragilis} prevented CDI morbidity in a mouse model and
inhibited \emph{C. difficile} adherence \emph{in vitro} (40). In
coculture experiments \emph{B. longum} decreased \emph{C. difficile}
biofilm formation while \emph{B. thetaiotamicron} enhanced biofilm
formation (41) and \emph{B. dorei} reduced \emph{C. difficile} growth in
a 9-species community \emph{in vitro} (42). Therefore, whether
\emph{Bacteroides} is detrimental or beneficial in the context of
\emph{C. difficile} infection or colonization is still unclear, but the
niche and interactions with other bacteria may contribute.

@@

Despite identifying an altered compositional profile that included
higher relative abundance of the \emph{C. difficile} sequence in the
mucosal tissues of mice treated with 5-day PEG compared to the
clindamycin group, we did not see a difference in histopathology scores
between the groups. One reason there was no difference could be the
\emph{C. difficile} strain used, \emph{C. difficile} 630 results in mild
histopathology summary scores in mice compared to VPI 10463 despite both
strains producing toxin in mice (51). Part of our hypothesis for why
there could have been increased histopathology scores in PEG-treated
mice was because PEG was previously shown to disrupt the mucus layer in
mice. However, recent studies demonstrated that broad spectrum
antibiotics can also disrupt the host mucosal barrier in mice (52, 53).
Future research is needed to tease out the interplay between medications
that influence the mucus layer and different strains of \emph{C.
difficile} in the context of CDIs.

It is more difficult to interpret what are findings mean in the context
of \emph{C. difficile} colonization resistance in human patients. The
main difficulty being that most hospitals recommend not performing
\emph{C. difficile} testing if the patient is currently taking a
laxative. This recommendation is in accordance with the Infectious
Diseases Society of America and Society for Healthcare Epidemiology of
America guidelines (54). The rationale behind the recommendation is that
patients taking laxatives may be asymptomatically colonized with
\emph{C. difficile}, resulting in unnecessary antibiotic treatment
(55--57). Furthermore, some studies identified laxatives as a risk
factor for developing CDIs or recurrent CDIs (58--60) and a recent study
found the proportion of severe CDIs was similar between patients taking
and not taking laxatives (61). However, there have also been some
studies that suggest laxatives are not a risk factor for developing CDIs
(62, 63). Although, it is unclear whether laxatives impact CDI
susceptibility in human paitents, it is clear that laxatives also have a
transient impact on the human microbiota (13, 64--67). Further studies
to examine the relationship between laxatives, \emph{C. difficile}
colonization, and CDIs are warranted.

@@

\textbf{\emph{C. difficile} challenge model.} Mice were challenged with
25 microliters of \emph{C. difficile} 630 spores at
10\textsuperscript{5} concentration, except for 1 experiment where the
concentration was 10\textsuperscript{3} (Fig. 5A). All mock challenged
mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper
pipette was used to administer the same challenge dose to mice via oral
gavage. Mice were weighed daily throughout the experiment and stool was
collected for quantifying \emph{C. difficile} CFU and 16S rRNA gene
sequencing. There were two groups of mice that received either a PBS or
fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal
microbiota transplant was prepared with stool samples collected from the
mice in the experiment prior to the start of any treatments. The stool
samples were transferred to an anaerobic chamber and diluted 1:10 in
reduced PBS and glycerol was added to make a 15\% glycerol solution. The
solution was then aliquoted into tubes and stored at -80°C until the day
of the gavage. An aliquot of both the FMT and PBS solutions were also
set aside in the -80°C for 16S rRNA gene sequencing. The day of the
gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute.
The supernatant was then transferred to a separate tube to prevent the
gavage needle from clogging with debris during gavage. The PBS solution
that was administered to the other group was also 15\% glycerol. Each
mouse was administered 100 microliters of either the FMT or PBS solution
via gavage. When we refer to mice that cleared \emph{C. difficile}, we
mean that no \emph{C. difficile} was detected in the first serial
dilution (limit of detection: 100 CFU). In some experiments, we
collected tissues for 16SrRNA gene sequencing, histopathology, or both.
For 16S rRNA gene sequencing, we collected small snips of cecum,
proximal colon, and distal colon tissues in microcentrifuge tubes, snap
froze in liquid nitrogen, and stored at -80°C. For histopathology, cecum
and colon tissues were placed into separate casettes, fixed, and then
submitted to McClinchey Histology Labs (Stockbridge, MI) for processing,
embedding, and hematoxylin and eosin (H\&E) staining.

>>>>

Diarrheal samples from patients taking laxatives are typically rejected
for \emph{Clostridiodes difficile} testing. However, there are
similarities between the bacterial communities from people with diarrhea
or \emph{C. difficile} infections (CDI) including lower diversity
compared to communities from healthy patients. This observation led us
to hypothesize that diarrhea may be an indicator of \emph{C. difficile}
susceptibility. We explored how osmotic laxatives disrupt the
microbiota's colonization resistance to \emph{C. difficile} by
administering a laxative to mice either before or after \emph{C.
difficile} challenge. Our findings suggest that osmotic laxatives
disrupt colonization resistance to \emph{C. difficile}, and prevent
clearance among mice already colonized with \emph{C. difficile}.
Considering that most hospitals recommend not performing \emph{C.
difficile} testing on patients taking laxatives and laxatives are used
when administering fecal microbiota transplants via colonoscopy to
patients with recurrent CDIs, further studies are needed to evaluate if
laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative samples from patients were
separated into two groups based on stool consistency, there were similar
microbiota features between samples from CDI patients and \emph{C.
difficile} negative patients with diarrhea compared to non-diarrheal
samples that were \emph{C. difficile} negative (12). The similar
community features between CDI patients and patients with diarrhea
included low alpha diversity and there were also only 6 bacterial taxa
had higher relative abundances in communities from CDI patients. These
results led to the hypothesis that bacterial communities from patients
experiencing diarrhea are susceptible to developing CDIs, regardless of
how they developed diarrhea.

Depending on the dose administered, osmotic laxatives can lead to
diarrhea and temporarily disrupt the human intestinal microbiota (13).
The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
in Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} colonization
(14--17). The mucus barrier is thought to mediate protection from CDIs
by protecting intestinal epithelial cells from the toxins produced by
\emph{C. difficile} (18, 19). Whether laxative administration results in
more severe CDIs in mice and how long mice remain colonized with
\emph{C. difficile} after challenge is unclear.

@@

Because time and treatment group influenced most of the variation
between communities, we next explored whether there were differences in
community diversity and composition between treatment groups. We
examined the alpha diversity dynamics by calculating the communities'
Shannon diversity. Although both clindamycin and PEG treatments
decreased diversity, the Shannon diversity was lower in the groups of
mice that received PEG treatment compared to those that received
clindamycin alone through thirty days post-challenge (Fig. 2C; Data Set
S1, sheet 2). We next identified the bacterial genera whose relative
abundances shifted after PEG treatment by comparing the baseline samples
of mice treated with only PEG to samples from the same mice one day
post-PEG-treatment. We found 18 genera whose relative abundances were
altered by PEG treatment (Data Set S1, sheet 3). The majority of the
bacterial relative abundances decreased after the PEG treatment, but the
relative abundance among members of the \emph{Enterobacteriaceae} and
\emph{Bacteroides} increased. The increase in \emph{Bacteroides}
relative abundance was unique to PEG treated mice, as the
\emph{Bacteroides} relative abundance actually decreased in clindamycin
treated mice (Fig. 2D). Finally, we identified the genera whose relative
abundance differed across treatment groups over multiple time points. Of
the 33 genera that were different between treatment groups, 24 genera
were different over multiple time points (Fig. 2E, Data Set S1, sheet
4). Thus, PEG had a significant impact on the fecal microbiota that was
maintained over time and was distinct from clindamycin treatment.

@@

One possible explanation for the prolonged \emph{C. difficile}
colonization in 5-day PEG treated mice, might be due to the bacteria's
persistence in the mucosal compartment. In fact, it has been
hypothesized that \emph{C. difficile} biofilms may serve as reservoirs
for recurrent infections (35) and \emph{C. difficile} biofilms in the
mucus layer were recently identified in patients as aggregates with
\emph{Fusobacterium nucleatum} (36). There was an interesting pattern of
increased \emph{Enterobacteriaceae}, \emph{Bacteroides}, and \emph{C.
difficile} in both the stool and mucosal communities of PEG-treated mice
suggesting a potential synergy. \emph{Bacteroides} has the potential to
degrade mucus and the osmotic laxative may have allowed
\emph{Bacteroides} to colonize the mucosal niche by degrading mucin
glycans with glycosyl hydrolases that are absent in \emph{C. difficile}
(37). \emph{Bacteroides} persistence in the mucosal tissue might also
have helped \emph{Enterobacteriaceae} to colonize the region, as a
synergy between mucus-degrading \emph{B. fragilis} and \emph{E. coli}
has previously been described (38). A separate study demonstrated
\emph{C. difficile} was present in the outer mucus layer and associated
with \emph{Enterobacteriaceae} and \emph{Bacteroidaceae} using
fluorescent in situ hybridization (FISH) staining (39). However,
protective roles for \emph{Bacteroides} have also been demonstrated. For
example, \emph{B. fragilis} prevented CDI morbidity in a mouse model and
inhibited \emph{C. difficile} adherence \emph{in vitro} (40). In
coculture experiments \emph{B. longum} decreased \emph{C. difficile}
biofilm formation while \emph{B. thetaiotamicron} enhanced biofilm
formation (41) and \emph{B. dorei} reduced \emph{C. difficile} growth in
a 9-species community \emph{in vitro} (42). Therefore, whether
\emph{Bacteroides} is detrimental or beneficial in the context of
\emph{C. difficile} infection or colonization is still unclear, but the
niche and interactions with other bacteria may contribute.

@@

Despite identifying an altered compositional profile that included
higher relative abundance of the \emph{C. difficile} sequence in the
mucosal tissues of mice treated with 5-day PEG compared to the
clindamycin group, we did not see a difference in histopathology scores
between the groups. One reason there was no difference could be the
\emph{C. difficile} strain used, \emph{C. difficile} 630 results in mild
histopathology summary scores in mice compared to VPI 10463 despite both
strains producing toxin in mice (51). Part of our hypothesis for why
there could have been increased histopathology scores in PEG-treated
mice was because PEG was previously shown to disrupt the mucus layer in
mice. However, recent studies demonstrated that broad spectrum
antibiotics can also disrupt the host mucosal barrier in mice (52, 53).
Further research is needed to tease out the interplay between
medications that influence the mucus layer and different strains of
\emph{C. difficile} in the context of CDIs.

It is more difficult to interpret what are findings mean in the context
of \emph{C. difficile} colonization resistance in human patients. The
main difficulty being that most hospitals recommend not performing
\emph{C. difficile} testing if the patient is currently taking a
laxative. This recommendation is in accordance with the Infectious
Diseases Society of America and Society for Healthcare Epidemiology of
America guidelines (54). The rationale behind the recommendation is that
patients taking laxatives may be asymptomatically colonized with
\emph{C. difficile}, resulting in unnecessary antibiotic treatment
(55--57). Furthermore, some studies identified laxatives as a risk
factor for developing CDIs or recurrent CDIs (58--60) and a recent study
found the proportion of severe CDIs was similar between patients taking
and not taking laxatives (61). However, there have also been some
studies that suggest laxatives are not a risk factor for developing CDIs
(62, 63). Although, it is unclear whether laxatives impact CDI
susceptibility in human paitents, it is clear that laxatives also have a
transient impact on the human microbiota (13, 64--67). Additional
studies to examine the relationship between laxatives, \emph{C.
difficile} colonization, and CDIs are warranted.

@@

\textbf{\emph{C. difficile} challenge model.} Mice were challenged with
25 microliters containing 10\textsuperscript{5} \emph{C. difficile} 630
spores, except for 1 experiment where the concentration was
10\textsuperscript{3} (Fig. 5A). All mock challenged mice received 25 ul
vehicle solution (Ultrapure water). A Dymax stepper pipette was used to
administer the same challenge dose to mice via oral gavage. Mice were
weighed daily throughout the experiment and stool was collected for
quantifying \emph{C. difficile} CFU and 16S rRNA gene sequencing. There
were two groups of mice that received either a PBS or fecal microbiota
transplant (FMT) gavage post-PEG treatment. The fecal microbiota
transplant was prepared with stool samples collected from the mice in
the experiment prior to the start of any treatments. The stool samples
were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS
and glycerol was added to make a 15\% glycerol solution. The solution
was then aliquoted into tubes and stored at -80°C until the day of the
gavage. An aliquot of both the FMT and PBS solutions were also set aside
in the -80°C for 16S rRNA gene sequencing. The day of the gavage,
aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The
supernatant was then transferred to a separate tube to prevent the
gavage needle from clogging with debris during gavage. The PBS solution
that was administered to the other group was also 15\% glycerol. Each
mouse was administered 100 microliters of either the FMT or PBS solution
via gavage. When we refer to mice that cleared \emph{C. difficile}, we
mean that no \emph{C. difficile} was detected in the first serial
dilution (limit of detection: 100 CFU). In some experiments, we
collected tissues for 16SrRNA gene sequencing, histopathology, or both.
For 16S rRNA gene sequencing, we collected small snips of cecum,
proximal colon, and distal colon tissues in microcentrifuge tubes, snap
froze in liquid nitrogen, and stored at -80°C. For histopathology, cecum
and colon tissues were placed into separate cassettes, fixed, and then
submitted to McClinchey Histology Labs (Stockbridge, MI) for processing,
embedding, and hematoxylin and eosin (H\&E) staining.
